Leaf traits drive differences in biomass partitioning among major plant functional types

### 2015-07-01 manuscript.Rnw
minor edits to text

\textbf{The partitioning of biomass into leaves and stems is one of the most uncertain  and influential components of  global vegetation models (GVMs), affecting the residence time and storage of carbon in ecosystems, and the response of net primary production to rising CO\textsubscript{2}\citep{dekauwe2014,friend2014,ise2010}. Although GVMs typically assume that the major woody plant functional types (PFTs) differ in biomass partitioning, empirical data have been lacking. Using a new global database on biomass partitioning from over 175 studies\citep{Falster2015}, we here show a strong effect of PFT on partitioning between stems and leaves. We show that major woody PFTs do not differ in the amount of leaf area supported, but rather the amount of leaf mass per unit above ground biomass. Moreover, we find that these differences relate to systematic shifts in leaf mass per area, a leading trait within the leaf economics spectrum \citep{wrightetal2004}. In addition, we show how a second trait, leaf area per stem basal area, explains additional variation in biomass partitioning among species; however this trait does not differ systematically among major woody PFTs. Combined, these results uncover fundamental rules in how plants are constructed and allow for systematic benchmarking of biomass partitioning routines in current GVMs.
}

@@

Perhaps the biggest challenge for GVMs is to capture the combined responses of the more than 250,000 plant species comprising the world's vegetation. While most plants have the same basic resource requirements and physiological function, large differences exist among species in the amount of energy invested in different tissues (leaves, stems, roots). The approach taken by most GVMs for dealing with this functional diversity is to consider only a few archetypal plant functional types (PFTs) \citep{harrisonetal2010}, assumed to differ in key physiological attributes. While GVMs assume differences in biomass partitioning between PFTs (Methods \ref{sec:gvmsummary}), these predictions are constrained by limited empirical data. Moreover, there is little consensus on how biomass partitioning and allocation (see Methods \ref{sec:partterm} for terminology) should be modelled in GVMs \citep{franklin2012,dekauwe2014,friend2014}. These shortcomings largely reflect the lack of suitable datasets of global scope with which models can be tested, constrained and compared \citep{wolf2011}.

@@

Moreover, we show that differences in LMF among woody PFTs arise due to differences in leaf morphology, but they converge in the amount of leaf area held on the plant (Fig. 2b-c). Previously, it was unclear whether PFTs with higher leaf biomass also had higher total leaf area. Leaf area is relevant because it drives total light interception and thus photosynthesis. Several classic studies have assumed, based only on differences in measurements of leaf mass, that gymnosperms carry more leaf area than angiosperms \citep{chabothicks1982, bond1989}. We found no evidence for this, instead we found that the three PFTs maintain similar leaf area per unit above-ground biomass (the `leaf area ratio`, LAR).

@@

Another key metric governing vegetation behavior is the amount of leaf maintained per unit cross-sectional area of stem ($A_S$), a quantity closely related to hydraulic supply capacity \cite{enquist2003}. This ratio is also important because plant leaf mass is frequently estimated from $A_S$ with allometric equations \cite{shinozaki1964b, chave2005}.  Mirroring the results for LMF, we found that the amount of leaf mass per unit stem area ($M_F/A_S$) differs among the three PFTs, that these differences are also proportional to LMA (Fig. 2d), and that PFTs are similarly invariant in the amount of leaf area per unit stem area ($A_F/A_S$) (Fig. 2e).

@@

The finding that climate does not appreciably affect biomass partitioning is good news for models that parameterize the globe by PFT. Yet, these results contrast those of a recent study where LMF was found to increase with MAT across the globe \citep{reich2014}. One explanation for these differences is that our measurements were taken on individual plants, whereas in the other study they were on whole stands. This suggests climatic effects on stand-level biomass partitioning occur primarily by altering the size-distribution and stand density (number of individuals per unit area), rather than partitioning within individual plants. With the caveats that there is still variation between species and studies unaccounted for (Fig. S5), and that coarse climate variables such as MAT and MAP may mask small-scale climate effects on biomass partitioning, our results strongly suggest the effects of PFT are stronger than any direct climate effect on biomass partitioning within individual plants (Tables \ref{table/varpartB} and \ref{table/varpartMAPMAT}).

We found that LMF differs predictably among PFTs, however there was considerable variation in all studied variables between species within PFTs (Fig. S5). It is thus pertinent to understand whether we can move beyond a PFT-based classification by considering other traits that affect biomass partitioning. To do this, we decomposed $M_F/A_S$ into LMA ($M_F/A_F$), and the ratio of leaf area to stem cross-sectional area ($A_F/A_S$) (Eq. \ref{eq:M2}).

@@

Overall, our results establish general patterns about plant construction and thus lay an empirical base against which models can be benchmarked. We found that PFTs differed in LMF, and this was proportional to LMA. It thus seems possible to integrate a key leaf trait with whole-plant modelling of biomass partitioning. A recent study compared allocation routines in a number of leading ecosystem models \citep{dekauwe2014}, and recommended constraining allocation by observed biomass fractions instead of using constant allocation fractions. Based on our findings, a first approximation within GVMs would be to assume leaf area to stem cross-sectional area a parameter that does not vary between PFTs, and vary biomass allocation accordingly. LMA, already a parameter in most GVMs, then gives the ratio of leaf biomass to stem area. Some models already incorporate a similar algorithm (see Methods~\ref{sec:gvmsummary}), while other algorithms may be tuned to yield similar patterns between PFTs as we have presented. In any case, the growing availability of large datasets on stand biomass and individual plant construction  \citep{Falster2015} suggest the time is ripe for rigorous benchmarking \citep{abramowitz_towards_2012, dekauwe2014} of GVMs against empirical data.

>>>>

\textbf{The partitioning of biomass into leaves and stems is one of the most uncertain  and influential components of  global vegetation models (GVMs), affecting the residence time and storage of carbon in ecosystems, and the response of net primary production to rising CO\textsubscript{2}\citep{dekauwe2014,friend2014,ise2010}. Although GVMs typically assume that the major woody plant functional types (PFTs) differ in biomass partitioning, empirical data have been lacking. Using a new global database on biomass partitioning from over 175 studies\citep{Falster2015}, we here show that partitioning between stems and leaves varies strongly between PFTs (as defined by deciduous vs. evergreen and gymnosperm vs. angiosperm). We show that major woody PFTs do not differ in the amount of leaf area supported, but rather the amount of leaf mass per unit above ground biomass. Moreover, we find that these differences relate to systematic shifts in leaf mass per area, a leading trait within the leaf economics spectrum \citep{wrightetal2004}. In addition, we show how a second trait, leaf area per stem basal area, explains additional variation in biomass partitioning among species; however this trait does not differ systematically among major woody PFTs. Combined, these results uncover fundamental rules in how plants are constructed and allow for systematic benchmarking of biomass partitioning routines in current GVMs.
}

@@

Perhaps the biggest challenge for GVMs is to capture the combined responses of the more than 250,000 plant species comprising the world's vegetation. While most plants have the same basic resource requirements and physiological function, large differences exist among species in the amount of energy invested in different tissues (leaves, stems, roots). The approach taken by most GVMs for dealing with this functional diversity is to consider only a few archetypal plant functional types (PFTs) \citep{harrisonetal2010}, assumed to differ in key physiological attributes. While GVMs assume or predict differences in biomass partitioning between PFTs (Methods \ref{sec:gvmsummary}), these differences are constrained by limited empirical data. Moreover, there is little consensus on how biomass partitioning and allocation (see Methods \ref{sec:partterm} for terminology) should be modelled in GVMs \citep{franklin2012,dekauwe2014,friend2014}. These shortcomings largely reflect the lack of suitable datasets of global scope with which models can be tested, constrained and compared \citep{wolf2011}.

@@

Moreover, we show that differences in LMF among woody PFTs arise due to differences in leaf morphology, but they converge in the amount of leaf area held on the plant (Fig. 2b-c). Previously, it was unclear whether PFTs with higher leaf biomass also had higher total leaf area. Leaf area is relevant because it drives total light interception and thus photosynthesis. Several classic studies have assumed, based only on differences in measurements of leaf mass, that gymnosperms carry more leaf area than angiosperms \citep{chabothicks1982, bond1989}. We found no evidence for this, instead we found that the three PFTs maintain similar leaf area per unit above-ground biomass (the `leaf area ratio`, LAR), despite large differences in the leaf mass fraction.

@@

Another key metric governing vegetation behavior is the amount of leaf maintained per unit cross-sectional area of stem ($A_S$), a quantity  related to hydraulic supply capacity \cite{enquist2003}. This ratio is also important because plant leaf mass is frequently estimated from $A_S$ with allometric equations \cite{shinozaki1964b, chave2005}.  Mirroring the results for LMF, we found that the amount of leaf mass per unit stem area ($M_F/A_S$) differs among the three PFTs, that these differences are also proportional to LMA (Fig. 2d), and that PFTs are similarly invariant in the amount of leaf area per unit stem area ($A_F/A_S$) (Fig. 2e).

@@

The finding that climate does not appreciably affect biomass partitioning contrast those of a recent study where LMF was found to increase with MAT across the globe \citep{reich2014}. One explanation for these differences is that our measurements were taken on individual plants, whereas in the other study they were on whole stands. This suggests climatic effects on stand-level biomass partitioning occur primarily by altering the size-distribution and stand density (number of individuals per unit area), rather than partitioning within individual plants. With the caveats that there is still variation between species and studies unaccounted for (Fig. S5), and that coarse climate variables such as MAT and MAP may mask small-scale climate effects on biomass partitioning, our results strongly suggest the effects of PFT are stronger than any direct climate effect on biomass partitioning within individual plants (Tables \ref{table/varpartB} and \ref{table/varpartMAPMAT}).

We found that LMF differs predictably among PFTs, however there was considerable variation in all studied variables between species within PFTs (Fig. S5). It is thus pertinent to understand whether we can improve on a PFT-based classification by including other traits that affect biomass partitioning. To do this, we decomposed $M_F/A_S$ into LMA ($M_F/A_F$), and the ratio of leaf area to stem cross-sectional area ($A_F/A_S$) (Eq. \ref{eq:M2}).

@@

Overall, our results establish general patterns about plant construction and thus lay an empirical base against which models can be benchmarked. We found that PFTs differed in LMF, and that LMF was proportional to LMA. It thus seems possible to integrate a key leaf trait with whole-plant modelling of biomass partitioning. A recent study compared allocation routines in a number of leading ecosystem models \citep{dekauwe2014}, and recommended constraining allocation by observed biomass fractions instead of using constant allocation fractions. Based on our findings, a first approximation within GVMs would be to assume leaf area to stem cross-sectional area a parameter that does not vary between PFTs, and vary biomass allocation accordingly. LMA, already a parameter in most GVMs, then gives the ratio of leaf biomass to stem area. Some models already incorporate a similar algorithm (see Methods~\ref{sec:gvmsummary}), while other algorithms may be tuned to yield similar patterns between PFTs as we have presented. In any case, the growing availability of large datasets on stand biomass and individual plant construction  \citep{Falster2015} suggest the time is ripe for rigorous benchmarking \citep{abramowitz_towards_2012, dekauwe2014} of GVMs against empirical data.
